---
title: "RNASeqQuant: an R package for RNA-Seq quantification"
author: "Yulong Niu, Ruben Garrido-Oter"
date: "`r Sys.Date()`"
bibliography: RSQref.bib
csl: nature.csl
output:
  BiocStyle::html_document:
    toc: true
  BiocStyle::pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{RNASeqQuant: an R package for RNA-Seq quantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r style, echo=FALSE, results='asis', message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

``` {r macro, echo=FALSE, results='hide', message=FALSE}
RNASeqQuant <- function() {"*[RNASeqQuant](https://github.com/YulongNiu/RNASeqQuant)*"}
Robject <- function(x){sub('%obj%', x, '<span style="background-color:#F0F0F0;color:#404040;font-family:\'Lucida Console\', monospace">%obj%</span>')}
Rclass <- function(x){sub('%obj%', x, '<span style="font-family:\'Times New Roman\', Times, serif;font-style: italic">%obj%</span>')}
```

# Overview

`r RNASeqQuant()` implements both expectation maximization and gradient descent for RNA-Seq quantification.

# Models

## Single species RNA-Seq

For single species RNA-Seq, if $T=\{t_1, t_2, \dots, t_K\}$ is the set of transcripts (with length $l_i$ for $t_i$) and the copy number of $t_i$ is $c_i$, we define $\rho_k=\frac{c_k}{\sum\limits_{t \in T}c_t} = \frac{c_k}{M}$ is the relative abundances of $t_i$, where $M$ is the total copy number, so that $\sum\limits_{k=1}^K \rho_k=1$.

[@bray2016near; @pachter2011models]

## Multiple Species RNA-Seq




# References
