---
title: "RNASeqQuant: an R package for RNA-Seq quantification"
author: 
- Yulong Niu
- Ruben Garrido-Oter
date: "`r Sys.Date()`"
bibliography: RSQref.bib
csl: nature.csl
output:
  BiocStyle::pdf_document:
    toc: true
    includes:
      in_header: header.tex
  BiocStyle::html_document:
    toc: true
vignette: >
  %\usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{RNASeqQuant: an R package for RNA-Seq quantification}
  %\VignetteEngine{knitr::knitr}
---

```{r style, echo=FALSE, results='asis', message=FALSE}
options(tinytex.verbose = TRUE)
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

``` {r macro, echo=FALSE, results='hide', message=FALSE}
RNASeqQuant <- function() {"*[RNASeqQuant](https://github.com/YulongNiu/RNASeqQuant)*"}
Robject <- function(x){sub('%obj%', x, '<span style="background-color:#F0F0F0;color:#404040;font-family:\'Lucida Console\', monospace">%obj%</span>')}
Rclass <- function(x){sub('%obj%', x, '<span style="font-family:\'Times New Roman\', Times, serif;font-style: italic">%obj%</span>')}
```

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

\newcommand{\Pro}{\mathrm{P}}
\newcommand{\tildel}{\widetilde{l_t}}
\newcommand{\tildelu}{\widetilde{l_u}}
\newcommand{\tildesv}{\widetilde{l_v}}

$$
\DeclareMathOperator*{\argmax}{arg\,max}
$$

# Overview

`r RNASeqQuant()` implements both expectation maximization and gradient descent for RNA-Seq quantification.

# Models

## EM model for single species

In single species, if $T=\{t_1, t_2, \dots, t_K\}$ is the set of transcripts (with length $l_i$ for $t_i$) and the copy number of $t_i$ is $c_i$, we define $\rho_k=\frac{c_k}{\sum\limits_{t \in T}c_t} = \frac{c_k}{M}$ is the relative abundances of $t_i$, where $M$ is the total copy number, so that $\sum\limits_{k=1}^K \rho_k=1$.

For a single species RNA-Seq data-set, let $F=\{f_1, f_2, \dots, f_N\}$ be the set of transcription fragments (reads) in a total number of $N=|F|$. Fragments mapped to the transcript $t_i$ composes the set $F_t \in F$ in a number of $X_t=|F_t|$. We assume that all fragments in $F$ have the same length $m$. In $t_i$, the number of position in which the fragment can start is $\tildel = l_i - m + 1$. $\tildel$ is also called *effective length* [@bray2016near; @pachter2011models].

After mapping, we can observe the alignment positions of $f_i$, which can be mapped to several transcripts, but we do not know the exact transcript. The estimated parameters is $\alpha=\{\alpha_1, \alpha_2, \dots, \alpha_K\}$. The probability that the fragment $f$ comes from the transcripts $t$ is:

$$
\begin{equation}
\begin{split}
\Pro(f \in t) &= \frac{\rho_t M \tildel}{\sum\limits_{k=1}^{K} \rho_k M \widetilde{l_k}} \\
&= \frac{\rho_t \tildel}{\sum\limits_{k=1}^{K} \rho_k \widetilde{l_k}} \\
&= \alpha_t
\end{split}
\label{eq:1}
\end{equation}
$$

If $f$ comes from $t$, the probability that $f$ mapped to a certain position of $t$ is:

$$
\begin{equation}
\begin{split}
\Pro(\mathrm{pos}|f \in t) = \frac{1}{\tildel}
\end{split}
\label{eq:2}
\end{equation}
$$

Combining $\eqref{eq:1}$ and $\eqref{eq:2}$, the probability that $f$ mapped to a certain position of $t$ is:

$$
\begin{equation}
\begin{split}
\Pro(\mathrm{pos}, f \in t) = \frac{\alpha_t}{\tildel}
\end{split}
\label{eq:3}
\end{equation}
$$

The logarithm of likelihood (LL) function is:

$$
\begin{equation}
\begin{split}
LL &= \sum\limits_{f \in F} \log \left(\Pro(\mathrm{pos}|\alpha)\ \right) \\
&= \sum\limits_{f \in F} \log \left( \sum_{k=1}^{K} \Pro(\mathrm{pos}, f \in t_k|\alpha) \right) \\
&= \sum\limits_{f \in F} \log \left( \sum_{k = 1}^{K} y_k \frac{\alpha_k}{\widetilde{l_k}} \right) \\
\end{split}
\label{eq:4}
\end{equation}
$$

where $y_k$ is the $\{0, 1\}$ indicator.

We use expectation–maximization (EM) algorithm to estimate $\alpha$. In the $n$ iteration, $\Pro(f \in t_1|\mathrm{pos}, \alpha^{(n)})$ equals to:

$$
\begin{equation}
\begin{split}
\Pro(f \in t_1|\mathrm{pos}, \alpha^{(n)}) &= \frac{\Pro(f \in t_1, \mathrm{pos}|\alpha^{(n)})}{\sum\limits_{k=1}^{K} \Pro(f \in t_k, \mathrm{pos}|\alpha^{(n)})} \\
&= \frac{\alpha_1^{(n)} \frac{y_1}{\widetilde{l_1}}}{\sum\limits_{k=1}^{K} \alpha_k^{(n)} \frac{y_k}{\widetilde{l_k}}} \\
&= \lambda_1
\end{split}
\label{eq:5}
\end{equation}
$$

where $\sum\limits_{k=1}^{K} \lambda_k = 1$.

$$
\begin{equation}
\begin{split}
\alpha^{(n+1)} &= \argmax_\alpha \left( 
\sum_{f \in F} \sum_{k=1}^{K} \Pro(f \in t_k|\mathrm{pos}, \alpha^{(n)})  \log \left( \Pro(f \in t_1, \mathrm{pos}|\alpha^{(n)}) \right) 
\right) \\
&= \argmax_\alpha \left( 
\sum_{f \in F} \sum_{k=1}^{K} \lambda_k \log\left(
\alpha_k \frac{y_k}{\widetilde{l_k}}
\right)
\right) \\
&= \argmax_\alpha \left(
\sum_{f \in F} \sum_{k=1}^{K}  \lambda_k \log\left(
\alpha_k
\right)
\right)
\end{split}
\label{eq:6}
\end{equation}
$$

The $n+1$ estimation of $\alpha$ is:

$$
\begin{equation*}
\begin{split}
\alpha_k^{(n+1)} &= \frac{\sum_\limits{f \in F} \lambda_k}{\sum\limits_{i=1}^{K} \sum\limits_{f \in F} \lambda_i} \\
&= \frac{\sum\limits_{f \in F} \lambda_k}{N}
,\ k=1, 2, \dots, K
\end{split}
\end{equation*}
$$

## EM model for multiple species

For a mix species RNA-Seq data-set, species $A$ and $B$ for example, $R=\{r_1, r_2, \dots, r_U\}$ is the set of transcripts (with length $l_u$ for $r_u$) in species $A$ and $S=\{s_1, s_2, \dots, s_V\}$ is the set of transcripts (with length $l_v$ for $s_v$) in species $B$. Total transcripts in species $A$ and $B$ composes the set $T=\{t_1, t_2, \dots, t_{U+V}\}$. 

The relative abundances in $A$ and $B$ are $\rho_u$，$\sum\limits_{u=1}^U \rho_u= 1$ and $\theta_v$, $\sum\limits_{v=1}^V \theta_v= 1$, respectively. Similarly, we define:

$$
\begin{equation}
\begin{split}
\Pro(f \in r_u) = \frac{\rho_u \tildelu}{\sum\limits_{r \in R} \rho_r \widetilde{l_r}} = \alpha_u \\
\Pro(\mathrm{pos}, f \in r_u) = \frac{\alpha_u}{\tildelu}
\end{split}
\label{eq:7}
\end{equation}
$$

$$
\begin{equation}
\begin{split}
\Pro(f \in s_v) = \frac{\theta_v \tildesv}{\sum\limits_{s \in S} \theta_s \widetilde{l_s}} = \beta_v \\
\Pro(\mathrm{pos}, f \in s_v) = \frac{\beta_v}{\tildesv}
\end{split}
\label{eq:8}
\end{equation}
$$

The probability of single mapped fragment is:

$$
\begin{equation}
\begin{split}
\Pro(\mathrm{pos}|\eta)  &= \Pro(\mathrm{pos}, f \in A|\eta) + \Pro(\mathrm{pos}, f \in B|\eta)\\
&= \Pro(\mathrm{pos} | f \in A, \eta) \Pro(f \in A) + \Pro(\mathrm{pos} | f \in B, \eta) \Pro(f \in B) \\
&= \sum_{i=1}^{U}m_i \frac{\alpha_i}{\widetilde{l_i}} \Pro(f \in A) + \sum_{j=U+1}^{U+V}m_j \frac{\beta_j}{\widetilde{l_j}} \Pro(f \in B) 
\end{split}
\label{eq:9}
\end{equation}
$$

Thus, the parameters will be estimated are:

$$
\begin{equation*}
\begin{split}
\eta=\{\alpha_1, \dots, \alpha_u, \beta_1, \dots, \beta_v,  \Pro(f \in A),  \Pro(f \in B)\}
\end{split}
\end{equation*}
$$

In $n+1$ iteration, the parameters in the expanded EM algorithm:

$$
\begin{equation}
\begin{split}
\lambda_i^{(n+1)} &= \frac{\alpha_i^{(n)} \frac{m_i}{\widetilde{l_i}} \Pro(f \in A)}{\sum\limits_{i=1}^{U}\alpha_i^{(n)} \frac{m_i}{\widetilde{l_i}} \Pro(f \in A) + \sum\limits_{j=U+1}^{U+V}\beta_j^{(n)} \frac{m_j}{\widetilde{l_j}} \Pro(f \in B)}\\
\lambda_j^{(n+1)} &= \frac{\beta_j^{(n)} \frac{m_j}{\widetilde{l_j}} \Pro(f \in A)}{\sum\limits_{i=1}^{U}\alpha_i^{(n)} \frac{m_i}{\widetilde{l_i}} \Pro(f \in A) + \sum\limits_{j=U+1}^{U+V}\beta_j^{(n)} \frac{m_j}{\widetilde{l_j}} \Pro(f \in B)}\\
\alpha_i^{(n+1)} &= \frac{\sum\limits_{f \in F} \lambda_i^{(n+1)}}{\sum\limits_{f \in F} \sum\limits_{i=1}^{U} \lambda_i^{(n+1)}} \\
\beta_j^{(n+1)} &= \frac{\sum\limits_{f \in F} \lambda_j^{(n+1)}}{\sum\limits_{f \in F} \sum\limits_{j=U+1}^{U+V} \lambda_j^{(n+1)}} \\
\Pro(f \in A)^{(n+1)} &= \frac{\sum\limits_{f \in F} \sum\limits_{i=1}^{U} \lambda_i^{(n+1)}}{\sum\limits_{f \in F} \sum\limits_{i=1}^{U} \lambda_i^{(n+1)} + \sum\limits_{f \in F} \sum\limits_{j=U+1}^{U+V} \lambda_j^{(n+1)}} = \frac{\sum\limits_{f \in F} \sum\limits_{i=1}^{U} \lambda_i^{(n+1)}}{N}\\
\Pro(f \in B)^{(n+1)} &= \frac{\sum\limits_{f \in F} \sum\limits_{j=U+1}^{U+V} \lambda_j^{(n+1)}}{\sum\limits_{f \in F} \sum\limits_{i=1}^{U} \lambda_i^{(n+1)} + \sum\limits_{f \in F} \sum\limits_{j=U+1}^{U+V} \lambda_j^{(n+1)}} = \frac{\sum\limits_{f \in F} \sum\limits_{j=U+1}^{U+V} \lambda_j^{(n+1)}}{N} 
\end{split}
\label{eq:10}
\end{equation}
$$


# References
